TacticToe: Learning the Game of Tactics

### 2018-01-21 tactictoe_jv.tex
minor

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
% are effective from 
% tactic-level proofs and what kind of formulas are the results of effective 
% steps, 
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
Experiments on the \holfour standard 
library show that TacticToe is able to re-prove 59\% of 7088 theorems in less 
than 10 seconds. This is to be compared with the 29\% proved by the ATP Eprover 
in auto-schedule mode with 128 lemmas predicted and translated by HolyHammer.

@@

%
%Techniques combining machine learning with translation to automated
%reasoning have recently become an important component of formal proof 
%assistants. 
%Such ``hammer'' techniques complement traditional proof
%assistant automation as implemented by tactics and decision
%procedures. n this paper we present a unified proof assistant
%automation approach which attempts to automate the selection of
%appropriate tactics and tactic-sequences combined with an optimized
%small-scale hammering approach.
%
%We implement the technique as a tactic-level automation for HOL4:
%TacticToe. It implements a modified A*-algorithm directly in HOL4 that
%explores different tactic-level proof paths, guiding their selection
%by learning from a large number of previous tactic-level
%proofs. Unlike the existing hammer methods, TacticToe avoids
%translation to FOL, working directly on the HOL level. By combining tactic 
%prediction and premise selection, TacticToe is able to re-prove 39\% of 7902 
%HOL4 theorems in 
%%less than
%5 seconds whereas the best single HOL(y)Hammer strategy solves 32\% in the 
%same 
%amount of time.
\todo{include latest results: 66.56 percent of 7123 theorems in 60 seconds and
do a similar experiment for hh with minimization}

@@

\section{Proof Search}\label{sec:proofsearch}
\todo
{
TG:
describe the prior policy, the evaluation, 
the node selection algorithm,
the node extension (+ case of solved goal) + backpropagation.
add description of end of the algorithm.
}
Despite the best efforts of the prediction algorithms, the selected tactic may 
not 
solve the current goal, proceed in the wrong direction
or even loop. For that reason, the prediction needs to be accompanied by a  
proof search mechanism that allows for backtracking and 
can choose which proof tree to extend next and in 
which direction.
Our search algorithm is an advanced tree search algorithm that relies on 
evaluation and policy functions to guide the search.

\todo{TG: add transition}
A proof is in general not a path but a tree. This means that our search space 
has two branching factors: 
the choice of a tactic, and the number of goals produced by each tactic.
The first one can be thought as a disjunction of tactics. Indeed, 
if there exists a tactic that solves a goal then it is solved. And 
conversely, the second branching factor is a conjunction of goals. All pending 
goals must be proven in order to prove the parent goal. For example, in the 
case of induction tactics the pending goals are the base case and the inductive 
case.

\todo{TG: add general overview of the algorithm and describe the plan}
If we suppose that each tactic produces at most one goal, then the standard 
monte carlo tree search algorithm applies. So we will first discuss how our 
algorithm works in this case. And then explain how we extend it to support 
tactics producing 
multiple goals.

@@

Supervised learning introduction.
\subsection{Prior tactic policy}

@@

The exploitation function $Exploitation$ is the average of the evaluation of 
all nodes descendant of $a_i$ (itself included) in the current search tree.   
\[Exploitation{a_i} = Average_{b \in Desc(a_i))} {PriorEvaluation(b)}\]

@@

@@

\begin{table}[]
\centering
\setlength{\tabcolsep}{3mm}
\begin{tabular}{@{}ccccc@{}}
\toprule
\phantom{ab} & {arith} & {real} & {compl} 
& {meas} \\
\midrule
\tactictoe & 74.6 & 67.3 & 77.3 & 28.2\\
\eprover & 50.8 & 67.6 & 63.4 & 10.8 \\
\midrule
\phantom{abc}  & {proba} & {list} & {sort} & {f\_map} \\
\midrule
\tactictoe & 37.3 & 71.4 & 58.4 & 73.6 \\
\eprover & 19.3 & 20.7 & 15.8 & 20.9 \\
\bottomrule
\end{tabular}
\caption{\label{theories}Percentage (\%) of re-proved theorems in the theories 
\texttt{arithmetic}, \texttt{real}, \texttt{complex}, \texttt{measure},  
\texttt{probability}, \texttt{list}, \texttt{sorting} and \texttt{finite\_map}. 
}
\end{table}

@@

Constant progress: 
  From 45 percent on the same dataset in 5 seconds 
       59 percent in december in 10 seconds
  Probably 65-70 percent en of January in 60 seconds.

@@

\section{Case study: comparison with original proofs}\todo{
  TG + RK: describe the two examples.
  RK: After implementation and small experiment is done (in about 1 week) 
      Test tactictoe find new examples on new goals and manual exploration mode.
      Give feedback.
}
%Investigating further the different qualities of \tactictoe, we study its 
%generated proof scripts on an example in \texttt{list} theory (see 
%Example~\ref{ex:casestudy}). 
%The theorem to be proven states the equivalence 
%between the fact that a number $n$ is greater than the length of a list 
%$\mathit{ls}$ with the 
%fact that dropping $n$ elements from this list returns an empty list.
%
%The human proof proceeds by induction on $n$ followed by solving both goals 
%using rewrite steps combined with an arithmetic decision procedure. 
%Both \tactictoe proofs (NH and SH) follow the general idea of reasoning by 
%induction but solve 
%the base case and the inductive case in a different way.
%The base case only needs rewriting using the global simpset in the 
%\tactictoe(NH)
%proof, which is simulated by a call to \metis in the (SH) proof.
%The inductive case should require an arithmetic decision procedure as hinted 
%by 
%the human 
%proof. This is achieved by rewriting using an arithmetic simpset in the second 
%proof. In the first proof however, a rewriting step and case splitting step 
%were used
%to arrive at a point where \metis calls succeed.
%The tactic proof produced by \tactictoe(NH) often looks better than the one 
%discovered by \tactictoe(SH) in that it does not involve \metis calls with a 
%large numbers of premises.

@@

Original \holfour proof script (2.03 milliseconds)
\begin{lstlisting}[language=SMLSmall,frame=tb]
PROVE_TAC[IS_GCD,DIVIDES_ANTISYM]
\end{lstlisting}

>>>>

We bridge this gap by implementing our tactical prover TacticToe on top 
of the HOL4 ITP. 
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
% are effective from 
% tactic-level proofs and what kind of formulas are the results of effective 
% steps, 
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths. 
Experiments on the \holfour standard 
library show that TacticToe is able to re-prove 66.6\% of 7123 theorems in less 
than 60 seconds. 
This is to be compared with the 42.8\% proved in the same amount of time by a 
a barebone version of TacticToe.

In the same condtion the ATP Eprover proves ..., when it is run in 
auto-schedule mode with 128 lemmas predicted and 
translated by HolyHammer.
\todo{to do eprover results running}

@@

%
%Techniques combining machine learning with translation to automated
%reasoning have recently become an important component of formal proof 
%assistants. 
%Such ``hammer'' techniques complement traditional proof
%assistant automation as implemented by tactics and decision
%procedures. n this paper we present a unified proof assistant
%automation approach which attempts to automate the selection of
%appropriate tactics and tactic-sequences combined with an optimized
%small-scale hammering approach.
%
%We implement the technique as a tactic-level automation for HOL4:
%TacticToe. It implements a modified A*-algorithm directly in HOL4 that
%explores different tactic-level proof paths, guiding their selection
%by learning from a large number of previous tactic-level
%proofs. Unlike the existing hammer methods, TacticToe avoids
%translation to FOL, working directly on the HOL level. By combining tactic 
%prediction and premise selection, TacticToe is able to re-prove 39\% of 7902 
%HOL4 theorems in 
%%less than
%5 seconds whereas the best single HOL(y)Hammer strategy solves 32\% in the 
%same 
%amount of time.

@@

\section{Proof Search}\label{sec:proofsearch}
\todo
{
TG:
describe the prior policy, the evaluation, 
the node selection algorithm,
the node extension (+ case of solved goal) + backpropagation.
add description of end of the algorithm.
}
Despite the best efforts of the prediction algorithms, the selected tactic may 
not solve the current goal, proceed in the wrong direction
or even loop. For that reason, the prediction needs to be accompanied by a  
proof search mechanism that allows for backtracking and 
can choose which proof tree to extend next and in which direction.
Our search algorithm is a monte carlo tree search (MCTS) algorithm that relies 
on prior evaluation and policy functions. Those priors are learned through 
direct supervised learning in the case of the policy and via the data 
accumulated during orthogonalization for the evaluation.

So, we first give a static description of the proof tree and its component.
Then we explain how we compute the evaluation and policy functions from learned 
data. And finally, we show how one MCTS step changes the proof tree data 
structures.

%A proof is in general not a path but a tree. This means that our search space 
%has two branching factors: 
%the choice of a tactic, and the number of goals produced by each tactic.
%The first one can be thought as a disjunction of tactics. Indeed, 
%if there exists a tactic that solves a goal then it is solved. And 
%conversely, the second branching factor is a conjunction of goals. All pending 
%goals must be proven in order to prove the parent goal. For example, in the 
%case of induction tactics the pending goals are the base case and the 
%inductive 
%case.

%If we suppose that each tactic produces at most one goal, then the standard 
%monte carlo tree search algorithm applies. So we will first discuss how our 
%algorithm works in this case. And then explain how we extend it to support 
%tactics producing 
%multiple goals.

@@

Definition of solved goals, open goals  and pending goals.
Supervised learning introduction.

%A proof is in general not a path but a tree. This means that our search space 
%has two branching factors: 
%the choice of a tactic, and the number of goals produced by each tactic.
%The first one can be thought as a disjunction of tactics. Indeed, 
%if there exists a tactic that solves a goal then it is solved. And 
%conversely, the second branching factor is a conjunction of goals. All pending 
%goals must be proven in order to prove the parent goal. For example, in the 
%case of induction tactics the pending goals are the base case and the 
%inductive 
%case.

\subsection{Prior tactic policy}

@@

The exploitation function $Exploitation$ is the average of the evaluation of 
all nodes descendant of $a_i$ (itself included) in the current search tree.   
\[Exploitation(a_i) = Average_{b \in Desc(a_i))} {PriorEvaluation(b)}\]

@@

10 seconds experiment:

@@

\begin{table}[]
\centering
\setlength{\tabcolsep}{3mm}
\begin{tabular}{@{}ccccc@{}}
\toprule
\phantom{ab} & {arith} & {real} & {compl} 
& {meas} \\
\midrule
\tactictoe & 74.6 & 67.3 & 77.3 & 28.2\\
\eprover & 50.8 & 67.6 & 63.4 & 10.8\\
\midrule
\phantom{abc}  & {proba} & {list} & {sort} & {f\_map} \\
\midrule
\tactictoe & 37.3 & 71.4 & 58.4 & 73.6 \\
\eprover & 19.3 & 20.7 & 15.8 & 20.9 \\
\bottomrule
\end{tabular}
\caption{\label{theories}Percentage (\%) of re-proved theorems in the theories 
\texttt{arithmetic}, \texttt{real}, \texttt{complex}, \texttt{measure},  
\texttt{probability}, \texttt{list}, \texttt{sorting} and \texttt{finite\_map}. 
}
\end{table}

@@

60 seconds experiment + 
adding theorems from the namespace and removing 
additionally saved theorems + 
recording proofs from Prove.

Constant progress: 
+45 percent in Septermber in 5 seconds.
+59 percent in December in 10 seconds.
+66.6 percent in January in 60 seconds.

@@

\section{Case study: comparison with original proofs}\todo{
  TG + RK: find describe examples.
  RK: After implementation and small experiment is done (in about 1 week) 
      Test tactictoe find new examples on new goals and manual exploration mode.
      Give feedback.
}
%Investigating further the different qualities of \tactictoe, we study its 
%generated proof scripts on an example in \texttt{list} theory (see 
%Example~\ref{ex:casestudy}). 
%The theorem to be proven states the equivalence 
%between the fact that a number $n$ is greater than the length of a list 
%$\mathit{ls}$ with the 
%fact that dropping $n$ elements from this list returns an empty list.
%
%The human proof proceeds by induction on $n$ followed by solving both goals 
%using rewrite steps combined with an arithmetic decision procedure. 
%Both \tactictoe proofs (NH and SH) follow the general idea of reasoning by 
%induction but solve 
%the base case and the inductive case in a different way.
%The base case only needs rewriting using the global simpset in the 
%\tactictoe(NH)
%proof, which is simulated by a call to \metis in the (SH) proof.
%The inductive case should require an arithmetic decision procedure as hinted 
%by 
%the human 
%proof. This is achieved by rewriting using an arithmetic simpset in the second 
%proof. In the first proof however, a rewriting step and case splitting step 
%were used
%to arrive at a point where \metis calls succeed.
%The tactic proof produced by \tactictoe(NH) often looks better than the one 
%discovered by \tactictoe(SH) in that it does not involve \metis calls with a 
%large numbers of premises.

@@

Original \holfour proof script
\begin{lstlisting}[language=SMLSmall,frame=tb]
PROVE_TAC[IS_GCD, DIVIDES_ANTISYM]
\end{lstlisting}
